Bounded Model Checking for Embedded Systems \\ (Position Paper)

### 2016-01-04 paper.tex
extension of some topics throughout th whole paper and general corrections

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Our dependency on the correct functioning of embedded systems is rapidly growing, mainly due to their wide range of applications, such as micro-grids, automotive device control ({\it e.g.}, airbag control, etc.), health care, surveillance, mobile devices, and consumer electronics. Their structures are becoming more and more complex and now require multi-core processors with scalable shared memory, in order to meet increasing computational power demands. As a consequence, reliability of embedded (distributed) software becomes a key issue during system development, which must be carefully addressed and assured. Normally, state-of-the-art verification methodologies for embedded systems generate test vectors (with constraints) and use assertion-based verification and high-level processor model, during simulation; however, other additional challenges have been raised: the need for meeting time and energy constraints, handling concurrent software, dealing with platform restrictions, evaluating implementation-structure choices, and supporting legacy designs (usually written in low-level languages). The present paper discuss challenges, problems, and recent advances to ensure correctness and timeliness regarding embedded systems. Reliability issues, in the development of micro-grids and cyber-physical systems, are then considered, as a prominent model checking application.
\end{abstract}

@@

Generally, embedded computer systems perform dedicated functions with high degree of reliability. They are used in a variety of sophisticated applications, which range from entertainment software, such as games and graphics animation, to safety-critical systems, such as nuclear reactors and automotive controllers~\cite{Kopetz11}. Embedded systems are ubiquitous, in modern day information systems, and are also becoming increasingly important in our society, especially in micro-grids, where reliability and carbon emission reduction are of paramount importance, and in cyber-physical systems, which demand short development cycles and high-level of reliability~\cite{leeCPS}. As a consequence, human life has also become more and more dependent on the services provided by this type of system and, in particular, their success depends both on the provided service relevance and quality.

Indeed, many current embedded systems become interesting solutions only if they can reliably
perform their target tasks, such as unmanned aerial vehicles (UAVs) and medical monitoring systems. For instance, UAVs are a trend on military missions, due to the absence of pilots; however, an incorrect plan execution may also cost human-lives, which is unacceptable. In addition, wrong diagnosis or condition-evaluation reports also have the potential to compromise patients' health, with serious consequences.

@@

State-of-the-art verification methodologies for embedded systems generate test vectors (with constraints), use assertion-based verification, and high-level processor models during simulation~\cite{Behrend15,Lettnin09}; however, embedded system verification also raises additional challenges, such as time and energy constraints, handling of concurrent software, platform restrictions, and legacy designs, which are usually developed in low-level languages. In summary, the main challenges regarding the verification of embedded systems lie on improving coverage, where more system's functions are verified, reducing verification time, {\it i.e.}, pruning the state-space exploration during verification, and incorporating system models, which allow specific checks regarding system behavior and not only code correctness. For instance, the first two aspects are of extreme relevance in micro-grids and cyber-physical systems, in order to ensure reliability, which is a key issue for (smart) cities, industries, and consumers, and the third is essential in systems that implement scientific models, such as digital filters and controllers, which present a behavior that is highly dependent on signal inputs and outputs and whose deployment may be heavily affected by hardware restrictions.
%In summary, the main challenge when verifying embedded systems lies on improving coverage, where more system's functions are verified, and reducing verification time, {\it i.e.}, pruning the state-space exploration during verification. These aspects are of extreme relevance in micro-grids to ensure reliability, which is a key issue for (smart) cities and consumers.

@@

This position paper tackles four major problems in computer-aided verification for embedded systems: (RP1) providing suitable encodings into the Satisfiability Modulo Theories (SMT) \cite{BarrettSST09}, in particular, extend background theories typically supported by SMT solvers to reason accurately and effectively about realistic embedded programs, (RP2) exploiting SMT techniques to leverage bounded model checking of multi-threaded software, in order to mitigate the state-explosion problem due to threads interleavings, (RP3) proving correctness and timeliness of embedded systems, by taking into account stringent constraints imposed by hardware, and (RP4) incorporating knowledge about system purpose and associated features, in order to detect system-level failures.

@@

Bounded Model Checking (BMC) techniques based on Boolean Satisfiability (SAT)~\cite{handbook09} or SMT~\cite{BarrettSST09} have been successfully applied to verify single- and multi-threaded programs and find subtle bugs in real (embedded systems) software~\cite{Clarke04,MerzFS12,CordeiroF11,Ivancic05,Cordeiro12}. The idea behind the BMC techniques is to check the violation of a given property at a predetermined depth, {\it i.e.}, given a transition system \textit{M}, a property $ \phi $, and a limit of iterations \textit{k}, BMC unfolds the system \textit{k} times and converts it into a Verification Condition (VC) $ \psi $, such that $\psi$ is \textit{satisfiable} if and only if $\phi$ has a counterexample of depth
less than or equal to \textit{k}.

@@

A feasible alternative to check properties in BMC is to prove that an invariant (assertion) is \textit{k}-inductive~\cite{EenS03,Sheera00}. 
The simplest form of \textit{k}-induction consists of two steps: the \textit{base-case} and the \textit{induction-step}. In the base-case, we aim to check that $\phi$ holds in all states reachable from an initial state within $k$ steps (we assume that $k \geq 0$) and in the induction-step, we aim to check that whenever $\phi$ holds in $k$ consecutive states $s_1,\ldots,s_k$, $\phi$ also holds in the next state $s_{k+1}$ of the system. In particular, the \textit{k}-induction algorithm was successfully applied to ensure that (restricted) C programs do not contain data races~\cite{Donaldson10,Kinductor} and do respect time constraints specified during the system design phase~\cite{EenS03}. Additionally, the \textit{k}-induction is a well-established technique in hardware verification, where it is easier to be applied, due to the monolithic transition relation present in such designs~\cite{EenS03,GrosseLD09,Sheera00}. Recently, Gadelha {\it et al.} and Beyer {\it et al.} presented a new algorithm for proving correctness of (a large set of) C programs by mathematical induction, in a completely automatic way ({\it i.e.}, users do not need to provide the loop invariant)~\cite{Gadelha15,Beyer15}. However, there is still little evidence, in the associated literature, that model checking of embedded software, using \textit{k}-induction (and invariants), can be efficiently exploited in embedded system verification~\cite{Rocha15,Brain15}.

@@

It is worth noticing that, currently, SMT-based BMC approaches check for code properties in real programs, which basically address programing language issues and general correctness, without taking into account target applications or system behavior. Such a statement is important, since, as already mentioned, many system features are being moved to software domain, which then claims for schemes that not only check if source code is correctly written, but also if it will respond properly in real environments or under external problems. For instance, the anti-lock braking system software of a car model can be bug free, but it may not work properly if a sensor is damaged.

Indeed, research in software verification is now taking such considerations into account and some schemes already use knowledge about the system to be verified and the underlying hardware. Recently, Ismail {\it et al.} \cite{dsv_spin2015}  proposed a verification tool for digital systems, called digital system verifier (DSVerifier), which is able to aid engineers to check for overflow, limit cycle, output error, timing, stability, and minimum phase, considering finite word-length effects. DSverifier is a useful test tool, which takes into account realization forms ({\it e.g.}, direct forms, transposed forms, etc.) and other implementation restrictions. Ultimately, if some requirements are not met with a given configuration (realization structure and hardware restrictions), the error reports provided DsVerifier may suggest another setup. Finally, DSVerifier has already been used for verifying digital filters and controllers, which even include UAVs.

@@

\textbf{(RP1)} Cordeiro {\it et al.} proposed the first SMT-based BMC for full C programs (ESBMC)~\cite{Cordeiro12}, which was later extended to C++98 programs~\cite{ECBS13}. \textbf{(RP2)} This SMT-based BMC approach was further developed to verify deadlock, data races, and atomicity violations in multi-threaded software based on POSIX and CUDA libraries~\cite{CordeiroF11,Pereira15}, considering partial-order reduction and state-hashing techniques to prune the state-space exploration~\cite{morse15}. These verification methodologies and respective tools were also able to find undiscovered bugs related to arithmetic overflow, buffer overflow, and invalid pointer in, standard benchmarks, which were confirmed by the benchmark's creators ({\it e.g.}, NOKIA, NEC, NXP, and VERISEC). \textbf{(RP3)} Novel approaches to model check embedded software using \textit{k}-induction and invariants were proposed and evaluated, demonstrating is effectiveness in some real-life embedded systems applications~\cite{Gadelha15,Rocha15}. \textbf{(RP4)} The proposed SMT-based context-BMC was also extended to verify overflow, limit cycle, time constraints, stability, and minimum phase in discrete-time systems, in particular, digital filters and controllers~\cite{dsv_spin2015,esbmc_controller,esbmc_filter} \textcolor{red}{depois mencionar explicitamente UAVS} and to specify system-level properties using linear-time temporal logic~\cite{JMorse15}. Lastly, yet importantly, ESBMC represents one of the most prominent BMC tools (for verifying embedded systems) as observed in the international software verification competitions~\cite{MorseCNF13,MorseRCN014}.

@@

This paper presented the main challenges to verify design correctness in embedded systems. In particular, it emphasizes that stringent constraints imposed by the hardware ({\it e.g.}, real-time, memory allocation, interrupts, and concurrency) must be considered during verification. Additionally, there is little evidence that model checking embedded software using \textit{k}-induction (and invariants), which
extends BMC-based approaches from falsification to verification, can be applied to formally verify the correctness and the timeliness of embedded systems. Given that the complexity of software has increased significantly in embedded products, there are still some (recent) advances to stress and cover exhaustively the state space of the system, in order to verify low-level properties that have to meet the application's deadline, access the memory region, handle concurrency, and control the hardware registers \textcolor{red}{falar sobre a incorporacao e conhecimento sobre o sistema}. As future work, the main goal of this research is to extend BMC as a design and verification tool for achieving a correct-by-construction embedded systems implementation; special attention will be given to cyber-physical systems and modern micro-grids, considering small-scale versions of a distributed system so that reliability and other system-level properties ({\it e.g.}, carbon emission reduction in smart cities) are amenable to automated verification.

@@

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Our dependency on the correct functioning of embedded systems is rapidly growing, mainly due to their wide range of applications, such as micro-grids, automotive device control ({\it e.g.}, airbag control, etc.), health care, surveillance, mobile devices, and consumer electronics. Their structures are becoming more and more complex and now require multi-core processors with scalable shared memory, in order to meet increasing computational power demands. As a consequence, reliability of embedded (distributed) software becomes a key issue during system development, which must be carefully addressed and assured. Normally, state-of-the-art verification methodologies for embedded systems generate test vectors (with constraints) and use assertion-based verification and high-level processor model, during simulation; however, other additional challenges have been raised: the need for meeting time and energy constraints, handling concurrent software, dealing with platform restrictions, evaluating implementation-structure choices, and supporting legacy designs (usually written in low-level languages). The present paper discusses challenges, problems, and recent advances to ensure correctness and timeliness regarding embedded systems. Reliability issues, in the development of micro-grids and cyber-physical systems, are then considered, as a prominent model checking application.
\end{abstract}

@@

Generally, embedded computer systems perform dedicated functions with high degree of reliability. They are used in a variety of sophisticated applications, which range from entertainment software, such as games and graphics animation, to safety-critical systems, such as nuclear reactors and automotive controllers~\cite{Kopetz11}. Embedded systems are ubiquitous, in modern day information systems, and are also becoming increasingly important in our society, especially in micro-grids, where reliability and carbon emission reduction are of paramount importance, and in cyber-physical systems (CPS), which demand short development cycles and again high-level of reliability~\cite{leeCPS}. As a consequence, human life has also become more and more dependent on the services provided by this type of system and, in particular, their success depends both on the provided service relevance and quality.

Indeed, many current embedded systems, such as unmanned aerial vehicles (UAVs) and medical monitoring systems, become interesting solutions only if they can reliably
perform their target tasks. For instance, UAVs are a trend on military missions, due to the absence of pilots; however, an incorrect plan execution may also cost human-lives, which is unacceptable. In addition, wrong disease diagnosis or condition-evaluation reports also have the potential to compromise patients' health, with serious consequences.

Besides, when physical interaction with the real world is needed, which happens in CPS, additional care must be taken, mainly when human action is directly replaced, as in vehicle driving. Regarding the latter, even human-in-the-loop feedback control can be employed \cite{munir}, which then raises deeper concerns related to the reliability of human behavior modeling and system implementation.

@@

State-of-the-art verification methodologies for embedded systems generate test vectors (with constraints), use assertion-based verification, and high-level processor models during simulation~\cite{Behrend15,Lettnin09}; however, embedded system verification also raises additional challenges, such as time and energy constraints, handling of concurrent software, platform restrictions, and legacy designs, which are usually developed in low-level languages. In summary, the main challenges regarding the verification of embedded systems lie on improving coverage, where more system's functions are verified, reducing verification time, {\it i.e.}, pruning the state-space exploration during verification, and incorporating system models, which allow specific checks regarding system behavior and not only code correctness. Indeed, the first two aspects are of extreme relevance in micro-grids and cyber-physical systems, in order to ensure reliability, which is a key issue for (smart) cities, industries, and consumers, and the third one is essential in systems that implement device models, such as digital filters and controllers, which present a behavior that is highly dependent on signal inputs and outputs and whose deployment may be heavily affected by hardware restrictions.
%In summary, the main challenge when verifying embedded systems lies on improving coverage, where more system's functions are verified, and reducing verification time, {\it i.e.}, pruning the state-space exploration during verification. These aspects are of extreme relevance in micro-grids to ensure reliability, which is a key issue for (smart) cities and consumers.

@@

This position paper tackles four major problems in computer-aided verification for embedded systems: (RP1) providing suitable encoding into the Satisfiability Modulo Theories (SMT) \cite{BarrettSST09}, which may extend the background theories typically supported by SMT solvers, with the goal of reasoning accurately and effectively about realistic embedded programs, (RP2) exploiting SMT techniques to leverage bounded model checking of multi-threaded software, in order to mitigate the state-explosion problem due to thread interleaving, (RP3) proving correctness and timeliness of embedded systems, by taking into account stringent constraints imposed by hardware, and (RP4) incorporating knowledge about system purpose and associated features, which aims to detect system-level failures.

@@

Bounded Model Checking (BMC) techniques based on Boolean Satisfiability (SAT)~\cite{handbook09} or SMT~\cite{BarrettSST09} have been successfully applied to verify single- and multi-threaded programs and find subtle bugs in real (embedded system) software~\cite{Clarke04,MerzFS12,CordeiroF11,Ivancic05,Cordeiro12}. The idea behind the BMC techniques is to check the violation of a given property at a predetermined depth, {\it i.e.}, given a transition system \textit{M}, a property $ \phi $, and a limit of iterations \textit{k}, BMC unfolds the system \textit{k} times and converts it into a Verification Condition $ \psi $, such that $\psi$ is \textit{satisfiable} if and only if $\phi$ has a counterexample of depth
less than or equal to \textit{k}.

@@

A feasible alternative to check for properties in BMC is to prove that an invariant (assertion) is \textit{k}-inductive~\cite{EenS03,Sheera00}. 
The simplest form of \textit{k}-induction consists of two steps: the \textit{base-case} and the \textit{induction-step}. In the former, the goal to check that $\phi$ holds in all states that are reachable from an initial state, within $k$ steps ($k \geq 0$ is assumed). The latter, in turn, aims to check that whenever $\phi$ holds, in $k$ consecutive states $s_1,\ldots,s_k$, $\phi$ also holds in the next system state $s_{k+1}$.

In particular, the \textit{k}-induction algorithm was successfully applied to ensure that (restricted) C programs do not contain data races~\cite{Donaldson10,Kinductor} and do respect time constraints, which were specified during the system design phase~\cite{EenS03}. Besides, \textit{k}-induction is a well-established technique in hardware verification, where it is easier to be applied, due to the monolithic transition relation present in such designs~\cite{EenS03,GrosseLD09,Sheera00}.

Recently, Gadelha {\it et al.} and Beyer {\it et al.} presented a \textcolor{red}{conforme conversamos, eh um novo mesmo?} new algorithm for proving correctness of (a large set of) C programs, by mathematical induction, in a completely automatic way ({\it i.e.}, users do not need to provide the loop invariant)~\cite{Gadelha15,Beyer15}. However, there is still little evidence, in the associated literature, that model checking of embedded software, using \textit{k}-induction (and invariants), can be efficiently exploited in embedded system verification~\cite{Rocha15,Brain15}.

@@

It is worth noticing that, currently, SMT-based BMC approaches check for code properties in real programs, which basically addresses programing language issues and general correctness, without taking into account target applications or system behavior. Such a statement is important, since, as already mentioned, many system features are being moved to software domain, which then claims for schemes that not only check if the source code is correctly written, but also if it will properly respond in real environments or under external problems. For instance, the anti-lock braking system software of a vehicle model can be bug free, but it may not work correctly if a sensor is damaged.

Indeed, research in software verification is now taking such considerations into account and some schemes already use knowledge about the system to be verified and the underlying hardware. Recently, Ismail {\it et al.} \cite{dsv_spin2015}  proposed a verification tool for digital systems, called digital system verifier (DSVerifier), which is able to aid engineers to check for overflow, limit cycle, output error, timing, stability, and minimum phase, considering finite word-length effects. Indeed, DSverifier is a useful test tool, which takes into account realization forms ({\it e.g.}, direct forms, transposed forms, etc.) and other implementation restrictions and, ultimately, if the system requirements are not met with a given configuration, an analysis of the provided error report may suggest another setup. %Finally, DSVerifier has already been used for verifying digital filters and controllers, which even include UAV applications.

@@

\textbf{(RP1)} Cordeiro {\it et al.} proposed the first SMT-based BMC for full C programs (ESBMC)~\cite{Cordeiro12}, which was later extended to C++98~\cite{ECBS13}. \textbf{(RP2)} This SMT-based BMC approach was further developed to verify deadlock, data races, and atomicity violations in multi-threaded software based on POSIX and CUDA libraries~\cite{CordeiroF11,Pereira15}, considering partial-order reduction and state-hashing techniques, in order to prune the state-space exploration~\cite{morse15}. Such verification methodologies and respective tools were also able to find undiscovered bugs related to arithmetic overflow, buffer overflow, and invalid pointer, in standard benchmarks, which were confirmed by the benchmark's creators ({\it e.g.}, NOKIA, NEC, NXP, and VERISEC). \textbf{(RP3)} Novel approaches to model check embedded software using \textit{k}-induction and invariants were proposed and evaluated, demonstrating is effectiveness in some real-life embedded system applications~\cite{Gadelha15,Rocha15}. \textbf{(RP4)} The proposed SMT-based context-BMC was also extended to verify overflow, limit cycle, time constraints, stability, and minimum phase in discrete-time systems, in particular, digital filters and controllers~\cite{dsv_spin2015,esbmc_controller,esbmc_filter}, and to specify system-level properties using linear-time temporal logic~\cite{JMorse15}. In particular, a specific UVA application was tackled, in order to verify its attitude controllers \textcolor{red}{que artigo do iury mencionar?}. Lastly, yet importantly, ESBMC represents one of the most prominent BMC tools (for verifying embedded systems), as observed in the international software verification competitions~\cite{MorseCNF13,MorseRCN014}.

@@

This paper presented the main challenges related to the verification of design correctness in embedded systems. In particular, it emphasizes that stringent constraints imposed by the underlying hardware ({\it e.g.}, real-time, memory allocation, interrupts, and concurrency), along with system behavior models, must be considered during verification. Additionally, there is little evidence that model checking embedded software using \textit{k}-induction (and invariants), which
extends BMC-based approaches from falsification to verification, can be applied to formally verify the correctness and the timeliness of embedded systems.

Given that software complexity has significantly increased in embedded products, there are still some (recent) advances to stress and exhaustively cover the system state space, in order to verify low-level properties that have to meet the application's deadline, access memory regions, handle concurrency, and control hardware registers. Besides, there is a trend towards incorporating knowledge about the system to be verified, which may take software verification to another step, where not only code correctness will be addressed, but also full system reliability.

As future work, the main goal of this research is to extend BMC as a design and verification tool for achieving a correct-by-construction embedded system implementations. Special attention will be given to cyber-physical systems and modern micro-grids, considering small-scale versions of a distributed system, so that reliability and other system-level properties ({\it e.g.}, carbon emission reduction in smart cities) are amenable to automated verification, probably through behavior models.

@@

\bibitem{munir}
Munir  S., Stankovic J. A., Liang C.-J. M., Lin S.:
\newblock {\em Cyber Physical System Challenges for Human-in-the-Loop Control}. 
\newblock In: 8th International Workshop on Feedback Computing, pp. 1--4, 2013.
